---
title: "Glass Ceilings? In our film industry? It's more likely than you think."
author: "Selina Ho, Beth Campbell, Olivia Baldwin"
date: "05/08/2018"
output:
  html_document:
    code_folding: hide
    df_print: paged
---

```{r message = FALSE, echo = FALSE}
library(tidyverse)
library(mdsr)
library(RMySQL)
db <- dbConnect_scidb(dbname = "imdb")
```
#Glass ceilings? In our film industry? It’s more likely than you think.

We’ve recently witnessed the downfall of some of the most iconic men in Hollywood: Harvey Weinstein, Louis C.K, Bill Cosby, Charlie Rose, Kevin Spacey, Mario Batali, George Takei, etc. But troubling questions still linger: what other institutions were complicit in their behavior? Who else had known? What happens to the machine behind the man in the aftermath? Even as Charlie Rose is reportedly planning his comeback [^1], more revelations about how long complaints have been lodged against him and ignored are being published. The earliest instance of harassment the Washington Post found happened, shockingly, over three decades ago [^2]. In 1986, Annmarie Parr’s concerns about Rose’s lewd comments were dismissed by an unnamed senior producer. Throughout the decades, executive producer after executive producer continued this same pattern of behavior, calling Rose “harmless.” All of these questions lead us to a larger conversation about gender inequality: the importance of not only scrutinizing individual acts of misogyny but also dismantling structural inequality that creates this culture of impunity and behavior. One of the most touted reasons for inaction and willful ignorance of these men’s behavior is assurance of continued profit, so we wanted to find out if there was a relationship between the profits of a production company and the gender disparity of the directors they finance[^3]. However, we also acknowledge there are victims and perpetrators of harassment of every gender. 

```{r message = FALSE, echo = FALSE}
query1 <- "SELECT t.title, t.production_year, mi.info, cn.name
FROM movie_info mi
LEFT JOIN title t ON mi.movie_id = t.id
LEFT JOIN movie_companies mc ON mi.id = mc.movie_id
LEFT JOIN company_name cn ON t.phonetic_code = cn.name_pcode_sf
WHERE kind_id = 1
AND company_type_id = 2
AND production_year >= 2000
GROUP BY cn.name
LIMIT 20;"
#db %>%
#  dbGetQuery(query1)
# language table
query3 <- "SELECT t.id, t.title, mi.info AS lang, t.production_year AS production_year, cn.name AS production_company_name
FROM title t
JOIN movie_info mi ON mi.movie_id = t.id
LEFT JOIN movie_companies mc ON mi.id = mc.movie_id
LEFT JOIN company_name cn ON t.phonetic_code = cn.name_pcode_sf
WHERE mi.info_type_id = 4 AND mi.info LIKE '%English%'
AND kind_id = 1
#AND company.type_id = 2
AND production_year >= 2000
GROUP BY cn.name
LIMIT 100;"
lang_table <- db %>%
  dbGetQuery(query3)
#lang_table

# movie's name, gross, director, gender of director, and its production company but limited by the year
query2 <- "SELECT t.title, t.id, mi.info AS movie_gross, n.name AS director, n.gender, co.name AS production_company
FROM cast_info ci
LEFT JOIN title t ON ci.movie_id = t.id
LEFT JOIN char_name cn ON cn.id = ci.person_role_id
LEFT JOIN name n ON n.id = ci.person_id
LEFT JOIN movie_info mi ON t.id = mi.movie_id
LEFT JOIN movie_companies mc ON mi.id = mc.movie_id
LEFT JOIN company_name co ON t.phonetic_code = co.name_pcode_sf
WHERE t.kind_id = 1
AND ci.role_id = 8
AND company_type_id = 2
AND info_type_id = 107
AND production_year >= 2000
AND mi.info LIKE '%Worldwide%' OR '%(USA)%'
GROUP BY ci.person_id;"

query4 <- "SELECT t.title, t.id, mi.info AS movie_budget, n.name AS director, n.gender, co.name AS production_company
FROM cast_info ci
LEFT JOIN title t ON ci.movie_id = t.id
LEFT JOIN char_name cn ON cn.id = ci.person_role_id
LEFT JOIN name n ON n.id = ci.person_id
LEFT JOIN movie_info mi ON t.id = mi.movie_id
LEFT JOIN movie_companies mc ON mi.id = mc.movie_id
LEFT JOIN company_name co ON t.phonetic_code = co.name_pcode_sf
WHERE t.kind_id = 1
AND ci.role_id = 8
AND company_type_id = 2
AND info_type_id = 105
AND production_year >= 2000
GROUP BY ci.person_id;"
# save table
director_gender_prod_comp <- db %>%
  dbGetQuery(query2)
# parse gross
 director_gender_prod_comp <- director_gender_prod_comp %>%
  filter(!is.na(gender)) %>%
  mutate(gross = parse_number(movie_gross))
 
 budget_plot <- db %>%
  dbGetQuery(query4)
 
budget_plot <- budget_plot %>%
  filter(!is.na(gender)) %>%
  mutate(budget = parse_number(movie_budget))

 #director_gender_prod_comp

# tab1 <- left_join(lang_table, select(director_gender_prod_comp, id, title, movie_gross, director, gender), by = 'id')
# 
# tab2 <- left_join(lang_table, select(director_gender_prod_comp, id, movie_gross, director, gender), by = 'id')

table_plot <- director_gender_prod_comp %>%
  mutate(is_femme = ifelse(gender == "f", 1, 0)) %>%
  group_by(production_company) %>%
  summarize(num_f = sum(is_femme),
            num_directors = count(!is.na(director)),
            sum_gross = sum(gross),
            percent_f_director = num_f/num_directors) %>%
  filter(sum_gross >= 300000000) %>%
  filter(!is.na(production_company)) %>%
  arrange(desc(percent_f_director))
table_plot 

table_plot %>%
  ggplot(aes(x = reorder(production_company,-sum_gross), y = sum_gross, fill = percent_f_director)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  scale_fill_gradient(high = "#032c6d", low = "#89b6ff",
                      breaks = c(0.00, 0.33, 0.66, 1.00),
                      labels = c("0%", "33.3%", "66.7%", "100%")) +
  theme(axis.text.x = element_blank()) +
  labs(title = "Total Gross of Production Companies vs Gender Disparity of Directors",
       x = "Production Companies",
       y = "Total Gross, 2000-2018 ($)",
       fill = "Percent of\nFemale Directors",
       caption = "Source: Internet Movie Database") +
  scale_y_continuous(breaks = c(0e+00, 2e+09, 4e+09, 6e+09, 8e+09),
                     labels = c("0", "2 billion", "4 billion", "6 billion", "8 billion"))

table_3_plot <- director_gender_prod_comp %>%
  mutate(is_femme = ifelse(gender == "f", 1, 0)) %>%
  group_by(gender) %>%
  summarize(num_films = n_distinct(title),
            sum_gross = sum(gross),
            avg_gross = sum_gross/num_films) %>%
  filter(!is.na(gender)) %>%
  arrange(desc(avg_gross))
table_3_plot

table_3_plot %>%
  ggplot(aes(x = reorder(gender,-avg_gross), y = avg_gross, fill = gender)) +
  geom_bar(stat = "identity", position = position_dodge())

table_2_plot <- budget_plot %>%
  mutate(is_femme = ifelse(gender == "f", 1, 0)) %>%
  group_by(gender) %>%
  summarize(num_films = n_distinct(title),
            sum_budget = sum(budget),
            avg_budget = sum_budget/num_films) %>%
  filter(!is.na(gender)) %>%
  arrange(desc(avg_budget))
table_2_plot

table_2_plot %>%
  ggplot(aes(x = reorder(gender,-avg_budget), y = avg_budget, fill = gender)) +
  geom_bar(stat = "identity", position = position_dodge())

```
Using the Internet Movie Database, we found movie production companies whose releases since 2000 grossed a total of over $300 million. And of those companies, we found what percentage of their directors are female. Across all gross amounts, female representation was extremely scarce. Over the span of 18 years, not one production company financed more than two female-helmed movies and the vast majority of companies financed zero. One company financed two, seven financed one, the other 84 companies financed zero female-directed movies. The production companies with the highest percentage of gender parity were predominantly earning the average gross profit with some companies in the upper echelons. 

Movie studios do no give behind-the-scene roles to women simply because it would be a deviation to the norm. Since Hollywood was created around 1919, men have been the primary moviemakers. Although women have made their way into the business, currently, it is still not 50/50 as it should be. Therefore, movie studios view giving women these jobs as a risk. Why give a woman a producing job when a white male is available? These movie studios have made millions in the past creating movies from the help of white men. At this point, movie studios can argue that the American public has a demand for movies made by white men and portraying white men. They see no reason to change. Rarely when movie studios do give the reins to women, they don’t receive as much financial payoff (see Figure 2). However, this is due to a variety of reasons and is not necessarily the woman director’s fault. Films directed by women receive less resources than films directed by men. These films are also not as widely distributed and therefore are much less likely to bring in money. Women made movies, however, bring in a “greater return on their investment” according to a study done by Slated in 2016 [^6]. Therefore, although movie studios can claim that it’s a risk to change their habits and hire more women, the data shows that this is most likely a cover for their deeply rooted misogyny instead of a concern for their finances.

The #metoo movement marked a turning point in the film industry and their fight against the systemic oppression of women. The movement was founded by the activist Tarana Burke in 2006 as a grassroots movement “to help survivors of sexual violence, particularly young women of color from low wealth communities, find pathways to healing” but became a viral international dialogue. After the downfall of Harvey Weinstein in fall of 2017, more and more women have been coming forward with accusation against powerful men in film and beyond. Although there is still a long way to go, Hollywood is witnessing a change in its culture. An example of this change comes from the Time’s Up initiative, created by around 300 Hollywood executives and actresses. On January 1, 2018, the initiative published a letter in The New York Times calling for legislation that would punish companies who allow harassment and gender equality in movie studios. Time’s Up also created a legal defense fund which would help underprivileged women defend themselves against harassment in their workplace. The Time’s Up movement is led by many powerful women in Hollywood including Shonda Rhimes, Ashley Judd, Eva Longoria, America Ferrera, Natalie Portman, Rashida Jones, Emma Stone, Kerry Washington and Reese Witherspoon [^5]. The cultural movement that Time’s Up and #metoo has sparked is seen through many recent and future films including A Wrinkle in Time, Ocean’s 8, Mudbound, Lady Bird, Wonder Woman, etc. These movies, either directed by a woman or featuring a female cast, have given women the financial platform to make films never quite seen before. Men still hold the most powerful positions in Hollywood and it will take years for that power dynamic to change. However, women have started to ask for more from movie studios. And men have realized that in order to be on the right side of history, they must give it to them.

[^1]:https://www.vanityfair.com/hollywood/2018/04/charlie-rose-tv-series-me-too-scandal-matt-lauer-louis-ck-tina-brown
[^2]:https://www.washingtonpost.com/charlie-roses-misconduct-was-widespread-at-cbs-and-three-managers-were-warned-investigation-finds/2018/05/02/80613d24-3228-11e8-94fa-32d48460b955_story.html?noredirect=on&utm_term=.3377d5638ca0
[^3]: https://github.com/obaldwin/mp4
[^4]: Miss Representation Trailer https://youtu.be/W2UZZV3xU6Q
[^5]: https://www.nytimes.com/2018/01/01/movies/times-up-hollywood-women-sexual-harassment.html
[^6]: https://www.hollywoodreporter.com/news/study-films-directed-by-women-907229
